Ricci curvature of the rooted phylogenetic subtree-prune-regraft graph

### 2015-02-10 curvature.tex
rough MH curvature bound

\begin{lemma}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	We first prove (1).
	Clearly $\degree{T} / \degree{S}$ is minimized when $T$ is of minimum degree and $S$ of maximum degree.
	Then, by Lemma~\ref{lem:degree_extremes}, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$.
	We thus have:
	\begin{align}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\
		\ge\ &\frac{3}{4} &\forall n \ge 3
	\end{align}

	We now prove (2).
	$\degree{S} - \degree{T}$  is maximized when $T$ and $S$ are of minimum and maximum degree.
	We thus have:
	\begin{align}
		\degree{S} - \degree{T}
		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
		- (3n^2 - 13n + 14) \\
		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		\le\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		\le\ &n^2 - 5n + 6
	\end{align}
\end{proof}

@@

	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is maximized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.
	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
	We then have:
	\begin{align}
		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align}
	Thus (2) and (3) follow.

@@

\begin{proof}
	We first prove the lower bound, that is $\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,s)$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\kappa_{l,0}(T,S) = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\kappa_{l,p}(T,S) = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
	Then:
	\begin{align}
		\kappa_{l,p}(T,S) &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&\ge \kappa_{l,0}(T,S)
	\end{align}

@@

>>>>

\begin{lemma}
	\label{lem:degree_max_delta}
	Let $T$ and $S$ be rooted phylogenetic trees with $n \ge 3$ leaves, and assume without loss of generality that $\degree{T} \le \degree{S}$.
	Then:
	\begin{enumerate}
		\item $\frac{\degree{T}}{\degree{S}} \ge \frac{3}{4}$, and
		\item $\degree{S} - \degree{T} \le n^2 - 5n + 6$.
	\end{enumerate}
\end{lemma}
\begin{proof}
	We first prove (1).
	Clearly $\degree{T} / \degree{S}$ is minimized when $T$ is of minimum degree and $S$ of maximum degree.
	Then, by Lemma~\ref{lem:degree_extremes}, $\degree{T} = 3n^2 - 13n + 14$ and $\degree{S} = 4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}$.
	We thus have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} \ge\ &\frac{3n^2 - 13n + 14}{4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)}} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4(n-2)^2 - 2(n-2)} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 16n + 16 - 2(n-2)} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 18n + 20} \\
		\ge\ &\frac{3n^2 - 13n + 12}{4n^2 - 17\frac{1}{3}n + 18} &\forall n \ge 3 \\
		\ge\ &\frac{3}{4} &\forall n \ge 3
	\end{align*}

	We now prove (2).
	$\degree{S} - \degree{T}$  is maximized when $T$ and $S$ are of minimum and maximum degree.
	We thus have:
	\begin{align*}
		\degree{S} - \degree{T}
		\le\ &(4(n-2)^2 - 2 \sum_{m=1}^{n-2} \floor{\log_2(m+1)})
		- (3n^2 - 13n + 14) \\
		\le\ &(4(n-2)^2 - 2(n-2)) - (3n^2 - 13n + 14) \\
		\le\ &4n^2 - 16n + 16 - 2n +4 - 3n^2 + 13n - 14 \\
		\le\ &n^2 - 5n + 6
	\end{align*}
\end{proof}

@@

	The relative change in degree, $\frac{\degree{T}}{\degree{S}}$, is maximized when $\degree{T}$ is as small as possible and $\degree{S} - \degree{T}$ is maximal.
	That is, $\degree{S} = 3n^2 - 13n + 14$ and $\degree{S} + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}$.
	We then have:
	\begin{align*}
		\frac{\degree{T}}{\degree{S}} &= \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + 2\floor{\frac{n-2}{2}} \ceil{\frac{n-2}{2}}} \\
		&\ge \frac{3n^2 - 13n + 14}{3n^2 - 13n + 14 + \frac{1}{2}(n-2)^2} \\
		&\ge \frac{3n^2 - 13n + 14}{3.5n^2 - 15n + 16}
	\end{align*}
	Thus (2) and (3) follow.

@@

\begin{proof}
	We first prove the lower bound, that is $\kappa_{l,0}(T,S) \le \kappa_{l,p}(T,s)$.
	Let $W_1(T,S)$ be the mass transport cost in the uniform case, and $W_1'(T,S)$ be the same for the lazy case with parameter $p$.
	Recall that $\kappa_{l,0}(T,S) = 1 - \frac{W_1(T,S)}{\dspr{T,S}}$, and $\kappa_{l,p}(T,S) = (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p$.
	Observe that $$W_1'(T,S) \le pW_1(T,S) + (1-p)\dspr{T,S},$$ by taking an optimal uniform mass assignment spread over $p$ of the mass and moving the remainder from $T$ to $S$ (or vice versa).
	Then:
	\begin{align*}
		\kappa_{l,p}(T,S) &= (1 - \frac{W_1'(T,S)}{\dspr{T,S}}) / p \\
		&\ge (1 - \frac{pW_1(T,S) + (1-p)\dspr{T,S}}{\dspr{T,S}}) / p \\
		&\ge (\frac{1}{p} - \frac{W_1(T,S)}{\dspr{T,S}} - \frac{1-p}{p}) \\
		&\ge 1 - \frac{W_1(T,S)}{\dspr{T,S}} \\
		&\ge \kappa_{l,0}(T,S)
	\end{align*}

@@

We continue to explore the effect of biased random walks by considering the curvature of Metropolis Hastings random walks.
The Metropolis Hastings (MH) random walk proposes a move from a tree $T$ to a neighbor tree $S$ uniformly at random and then accepts the move according to the Hastings ratio, $\min\left(1, \frac{\degree{T}}{\degree{S}}\right)$.
The mass distribution for the MH random walk thus leaves a portion of mass at the origin tree, proportional to the relative degree difference of its higher degree neighbors.
We now bound the difference between the curvature of uniform random walks $\kappa$, and that of MH random walks, $\kappa_{MH}$.

\begin{lemma}
	Let $T$ and $S$ be phylogenetic trees with $n$ leaves. Then:
	\begin{enumerate}
	$$\kappa(T,S) - \frac{1}{3\dspr{T,S}}
	\le \kappa_{MH}(T,S)
	\le \kappa(T,S) + \frac{1}{3\dspr{T,S}}, \text{ and }$$ 
	$$\kappa(T,S) - \frac{1}{6}
	\le \kappa_{MH}(T,S)
	\le \kappa(T,S) + \frac{1}{6}.$$
\end{lemma}
\begin{proof}
	We first prove the lower bound.
	By Lemma~\ref{lem:degree_max_delta_adjacent}, the quotient of degrees for two adjacent trees $\ge \frac{5}{6}$.
	Thus, the hastings ratio is always $\ge \frac{5}{6}$.
	This implies that at most $\frac{1}{6}$ of the mass remains at tree $T$ in the mass distribution.
	Let $W_1(T,S)$ be the cost of an optimal mass transport for the uniform random walk from $T$ to $S$.
	We construct an upper bound on $W_1'(T,S)$ for the MH random walk by moving mass according to $W_1$ where possible, and moving the remainder either from $T$ to $S$, $T$ to a neighbor of $S$, or from $T's$ neighbors to $S$.
	The maximum possible mass that is not moved according to $W_1$ is $\frac{1}{6}$.
	Moreover, the affected mass must be moved through at most two additional trees.
	Then, $W_1' &\le W_1 + \frac{2}{6}$.
	We now have:
	\begin{align*}
		\kappa_{MH}(T,S) & \ge 1 - \frac{W_1 + \frac{1}{3}}{\dspr{T,S}} \\
		& \ge \kappa(T,S) - \frac{1}{3\dspr{T,S}}
	\end{align*}

	In the case that $\dspr{T,S}=1$, the affected mass must be moved through only at most one additonal tree, as $T$ and $S$ are adjacent.
	We thus obtain the lower bound of $\kappa(T,S) - \frac{1}{6}$ in this case.

	We obtain the upper bounds similarly to the lower bounds, by observing that the affected at most $\frac{1}{6}$ of the mass may move through at most two fewer trees (i.e. directly between $T$ and $S$ rather than a pair of neighbors at distance $\dspr{T,S} + 2$ from each other).
	Again, this is at most one fewer tree when $\dspr{T,S}=1$.
